\bf Visualising How Non-linear Dimension Reduction Warps Your Data

### 2024-01-11 paper.tex
fix gitignore so that questions of overwriting are not asked on git pull

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, enhanced, breakable, interior hidden, frame hidden, borderline west={3pt}{0pt}{shadecolor}, sharp corners]}{\end{tcolorbox}}\fi

@@

\hypertarget{non-linear-dimension-reduction-techniques}{%
\subsubsection{Non-linear dimension reduction
techniques}\label{non-linear-dimension-reduction-techniques}}

Non-linear dimension reduction techniques play a crucial role in the
analysis and visualization of high-dimensional data, where the
complexities of relationships among variables may not be adequately
captured by linear methods. Among these techniques, tSNE stands out for
its emphasis on preserving pairwise distances. By minimizing the
divergence between probability distributions in both the high and
low-dimensional spaces, t-SNE effectively reveals intricate structures
and patterns within the data. Its application is widespread in tasks
requiring the visualization of clusters and local relationships, though
it does require careful consideration of the perplexity parameter for
optimal results.

UMAP is another powerful non-linear technique that strikes a balance
between preserving local and global structures. Constructing a fuzzy
topological representation using a weighted k-nearest neighbors graph,
UMAP optimizes the low-dimensional embedding to resemble this
representation. Known for its efficiency and scalability, UMAP is
versatile across various scales of relationships in the data, although
parameter sensitivity, particularly concerning the choice of neighbors,
must be taken into account.

For trajectory data, PHATE provides specialized capabilities. It models
the affinity between data points, simulating a heat diffusion process to
capture developmental processes, particularly in single-cell genomics.
While PHATE excels in revealing trajectory structures and offering
insights into cellular development, it necessitates careful parameter
tuning due to its specialized nature.

TriMAP adopts a unique approach by approximating the data manifold
through the construction of a triangulated graph representation. This
technique efficiently captures both global and local structures by
representing the data as a network of triangles. TriMAP's strength lies
in its ability to efficiently capture complex structures, albeit with
sensitivity to parameter choices, including the number of neighbors.

In contrast, PaCMAP introduces supervised learning to create a
low-dimensional representation while preserving pair-wise relationships.
Constructing a graph based on pair-wise distances, PaCMAP optimizes an
embedding using a customizable loss function. Particularly notable is
PaCMAP's flexibility in incorporating class labels or additional
information to guide the embedding process, offering users a means to
customize its behavior and performance.

@@

\caption{\label{fig-nldervis}2D layouts from different NLDR techniques
applied the same data: (a) tSNE (perplexity = 27), (b) UMAP
(n\_neighbors = 50), (c) PHATE (knn = 5), (d) TriMAP (n\_inliers = 5,
n\_outliers = 4, n\_random = 3), and (e) PaCMAP (n\_neighbors = 10, init
= random, MN\_ratio = 0.9, FP\_ratio = 2). Is there a best
representation of the original data or are they all providing equivalent
information?}

@@

\caption{\label{fig-binsize}Hexbin plots from different number of bins
for the same \textbf{s\_curve\_noise\_umap} data: (a) b = (4, 8), s =
1.643542, (b) b = (12, 22), s = 1.643542, and (c) b = (21, 40), s =
1.643542. The hexbins are colored based on the density of points, with
yellow indicating higher point density and darker colors representing
lower point density within each bin. Does a value of number of bins
exist to effectively represent the low-dimensional data?}

@@

In our model tuning process, we strategically adjust three key
parameters to optimize the performance and accuracy of our approach.
They are (i) the total number of bins (\(b\)), (ii) a benchmark value to
remove low-density hexagons, and (iii) a benchmark value to remove long
edges.

\hypertarget{total-number-of-bins}{%
\subsection{Total number of bins}\label{total-number-of-bins}}

The total number of bins is a crucial parameter that defines the
granularity of our hexagonal grid. This parameter particularly
influential in capturing the intricacies of the data structure. It is
determined by multiplying the number of bins along the x-axis ((b\_1))
with the number of bins along the y-axis ((b\_2)), as per the formula:

\begin{equation}\protect\hypertarget{eq-equationp1}{}{
b = b_1 \times b_2
}\label{eq-equationp1}\end{equation}

Here, (b) represents the total number of bins. Adjusting the parameter
(b\_1) provides control over the total bin count ((b)), allowing us to
fine-tune the grid's configuration along the x-axis based on specific
requirements.

\hypertarget{benchmark-value-to-remove-low-density-hexagons}{%
\subsection{Benchmark value to remove low-density
hexagons}\label{benchmark-value-to-remove-low-density-hexagons}}

Addressing low-density hexagons is a systematic process to handle
sparsely represented data in certain regions. For each hex bin, we
identify the six nearest hex bins using an equal 2D distance metric.
Then, we calculate the mean density, as outlined in the equations:

\begin{equation}\protect\hypertarget{eq-equationp2}{}{
\text{standard count} = \frac{\text{count}}{\text{max count}} 
}\label{eq-equationp2}\end{equation}

\begin{equation}\protect\hypertarget{eq-equationp3}{}{
\text{mean density} = \frac{\text{standard count}}{6} 
}\label{eq-equationp3}\end{equation}

The standard count is derived from the number of observations in the hex
bins. By examining the distribution of mean densities and designating
the first quartile as the benchmark value, hex bins with mean densities
below this benchmark are removed. This process ensures the elimination
of regions with insufficient data density, focusing on areas with more
significant data representation and preserving the overall structure in
the low-dimensional space.

\hypertarget{benchmark-value-to-remove-long-edges}{%
\subsection{Benchmark value to remove long
edges}\label{benchmark-value-to-remove-long-edges}}

The removal of long edges is a critical step to create a smoother
representation by iteratively eliminating hexagons with excessive
distances between centroids. This process eliminates outliers and noise
while preserving essential local relationships within the data. To
achieve this, distances between vertices are sorted, and unique distance
values are extracted. By setting a threshold based on the largest
difference between consecutive distance values, long edges are
identified and removed. This refinement step contributes to enhancing
the quality of the triangular mesh, ensuring a more accurate
representation of the data structure.

@@

@@

{\centering \includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-diagnosticpltDiffParam-1.pdf}

\caption{\label{fig-diagnosticpltDiffParam}The .}

\hypertarget{sec-simpleex}{%
\subsection{Simulated data example}\label{sec-simpleex}}

In this section, we showcase the effectiveness of our methodology using
simulated data. The dataset comprises five spherical Gaussian clusters
in 4-\(d\), with each cluster containing an equal number of points and
consistent within variation.

We \emph{strongly} recommend viewing the linked videos for each study
while reading. Links to the videos are available in the figures for each
example. The videos show the visual appearance of the
\textbf{langevitour} interface with low-dimensional view and how we can
interact with the tour via the controls.

{\centering \includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-nldervis5Gau-1.pdf}

\caption{\label{fig-nldervis5Gau}2D layouts from different NLDR
techniques applied the same data: (a) tSNE (perplexity = 61), (b) UMAP
(n\_neighbors = 15), (c) PHATE (knn = 5), (d) TriMAP (n\_inliers = 5,
n\_outliers = 4, n\_random = 3), and (e) PaCMAP (n\_neighbors = 10, init
= random, MN\_ratio = 0.9, FP\_ratio = 2). Is there a best
representation of the original data or are they all providing equivalent
information?}

\includegraphics{paper_files/figure-pdf/unnamed-chunk-43-1.pdf}

@@

{\centering \includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-diagnosticpltPBMC-1.pdf}

\caption{\label{fig-diagnosticpltPBMC}The .}

@@

{\centering \includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-diagnosticpltZEI-1.pdf}

\caption{\label{fig-diagnosticpltZEI}The .}

>>>>

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, interior hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, sharp corners, boxrule=0pt, breakable]}{\end{tcolorbox}}\fi

@@

\hypertarget{introduction-to-umap}{%
\subsubsection{Introduction to UMAP}\label{introduction-to-umap}}

The UMAP algorithm constructs a fuzzy topological representation of the
high-dimensional data and then optimizes the low-dimensional embedding
to be as similar to this representation as possible \citep{Leland2018}.

The UMAP algorithm constructs a weighted k-nearest neighbors graph based
on the local similarity of data points in the original high-dimensional
space. The weighted graph is defined by a conditional probability
distribution \(P_{ij}\), which represents the probability of observing
data point \(x_i\) given data point \(x_j\) as a neighbor:

\[
P_{ij} = \frac{\exp(-\lVert x_i - x_j \rVert^2)}{\sum_{k \neq i} \exp(-\lVert x_i - x_k \rVert^2)}
\]

Similarly, a fuzzy topological representation of the data is constructed
in the low-dimensional space with conditional probability distribution
\(Q_{ij}\):

\[
Q_{ij} = \frac{\exp(-\lVert y_i - y_j \rVert^2)}{\sum_{k \neq i} \exp(-\lVert y_i - y_k \rVert^2)}
\]

where \(y_i\) and \(y_j\) are the corresponding low-dimensional
representations of data points \(x_i\) and \(x_j\).

UMAP minimizes the cross-entropy between the high-dimensional fuzzy
topological representation \(P_{ij}\) and the low-dimensional
representation \(Q_{ij}\) to find an optimal low-dimensional embedding.
The loss function for UMAP is defined as:

\[
L = \sum_{i} \sum_{j} P_{ij} \log \frac{P_{ij}}{Q_{ij}}
\]

The algorithm uses stochastic gradient descent to optimize the positions
of the data points in the low-dimensional space to minimize this loss
function.

In R, the \texttt{umap} package is often used to implement UMAP and
generate results.

The ``n\_neighbors'' parameter in UMAP determines the number of nearest
neighbors to consider for each data point during the construction of the
fuzzy topological representation. Higher values of ``n\_neighbors''
capture more global structure, while lower values emphasize local
structure. Choosing an appropriate value for ``n\_neighbors'' is
crucial, as it affects the balance between local and global structure in
the resulting low-dimensional representation.

@@

\caption{\label{fig-nldervis}2D layouts from different NLDR techniques
applied the same data: (a) tSNE (perplexity = 32), (b) UMAP
(n\_neighbors = 50), (c) PHATE (knn = 5), (d) TriMAP (n\_inliers = 5,
n\_outliers = 4, n\_random = 3), and (e) PaCMAP (n\_neighbors = 10, init
= random, MN\_ratio = 0.9, FP\_ratio = 2). Is there a best
representation of the original data or are they all providing equivalent
information?}

@@

\caption{\label{fig-binsize}Hexbin plots from different number of bins
for the same \textbf{s\_curve\_noise\_umap} data: (a) b = (4, 8), s =
2.031, (b) b = (9, 20), s = 2.031, and (c) b = (16, 36), s = 2.031. The
hexbins are colored based on the density of points, with yellow
indicating higher point density and darker colors representing lower
point density within each bin. Does a value of number of bins exist to
effectively represent the low-dimensional data?}

@@

@@

{\centering \includegraphics{paper_files/figure-pdf/fig-predict-1.pdf}

\caption{\label{fig-predict}The .}

@@

\hypertarget{sec-simpleex}{%
\subsection{Simulated data examples}\label{sec-simpleex}}

@@

@@
